The evolution of badges of status with learners

### 2019-07-02 QD_report.Rmd
FInalize QD report. Plot of evol simulations

---
title: "The evolution of badges of status with learners"
author: "Andrés Quiñones"
bibliography: SocialMorphCue.bib
output: 
  pdf_document: 
    fig_caption: yes
    keep_tex: yes
header-includes: \usepackage{float}
                \floatplacement{figure}{H}
---

@@

Classical Hawk-Dove game. Individuals have one fo two genetically determined phenotypic strategies. _Hawks_ are willing to start a conflict over resources, while _doves_ prefer to stand down in the hope to share the resource without an aggressive contest. Individuals interact randomly with each other over their lifetimes and collect resources depending on the strategy of theirs and their partners. In contrast to the classical _Hawk-dove_, in here when to _hawks_ interact with each other one of them wins the contest and gets $V-\frac{C}{2}$ as payoff; while the loser gets $-\frac{C}{2}$. Where $V$ is the value of the resource individuals are competing for, and $C$ is the cost of an agressive contest. individuals share the cost, but only the winner gets the value of the resource.The rest of payoff matrix follows the standard game.

@@

where $Q_A$ and $Q_b$ are the qualities of individuals A and B respectively, and $\beta$ is how important quality is in determining who wins the contest. 
Individuals' reproductive success is proportional to the payoff accumulated
throughout their lifes. Figure \ref{fig:HD_game} shows the evolutionary dynamics 
of  the two genotypes, which fit the game theoretical prediction.

@@

To complete the receiver side of the signalling system, we introduce a new 
strategy in the game that we call _learner_. _Learners_ use a mixed strategy in 
the hawk-dove game. Which means, they play hawk (dove) with a certain probability.
Furthermore, the probability with which the play each strategy changes during their lifetime according to the payoff they recieve in previous interactions. In other words, _learners_ learn to behave in the game. To implement the learning
process, we use the reinforcement learning formalism. Specifically, we use the actor-critic algorithm [@sutton_reinforcement_2018]. Individuals ( _learners_ ) perceive the _badge_ of their counterpart as the environmental state in which they find themselves. They estimate the value of that state and a preference for playing hawk (dove) in that state. They update their estimation and preference after each interaction they have. Given that the badge size is a real number between 0 and 1, there are infinitely many environmental states. Thus, for their value estimation and preference computation, individual must use a form of generalization. We let them
use as generalization a function approximation method based on radial basis functions. Each individual uses five _centers_ spread uniformly throughout the range of badge size variation (from 0 to 1). Each time an individual interacts with a conspecific the size of the badge of the interacting partner triggers are response from each one of the _centers_. This response decreases in strength as the badge size is further from the _center_. The decrease follows a radial basis function with gaussian shape. The sum of the responses given by all the centers provide the value estimation and the preference for one of the two pure strategies (hawk or dove). During the learning process individuals update the _weight_ of the response that each one of these centers triggers. In the case of the preference for the strategies the numerical value resulting from the sum of the responses is translated to a probability of one of the two pure strategies through a logistic function. Figure \ref{fig:learning_cartoonRBF} shows a made up example of how the function approximation works for the actor (top panel - probability of choosing dove) and the critic (bottom panel - expected payoff from facing an individual with a given badge size).

@@

In Figure \ref{fig:learning_invar} we show the phenotypic frequencies and interaction-type frequencies in 4 generations of a population composed of only _learners_ that do not vary in their quality. The population is also monomorphic for the reaction norm parameters ($\alpha=0$ and $\gamma=0$). With the values chosen the badge size distribution is the same as the quality distribution. Thus, the four generations respresent just replicates of the learning process. The results show that under quality invariance, the learning algorithm develops similar phenotypic frequencies as the evolutionary dynamics of the pure strategies.

@@

In figure \ref{fig:learning_var0.1} we show results of phenotypic and interaction types frequencies from simulations in which individuals vary in their quality. 
The value of quality is drawn from a truncated normal distribution with mean 0.5 
and standard deviation 0.1. Compared to the expectation from evolutionary dynamics of the standard game, the learning scenario with qualiy variation shows lower levels of hawks and therefore lower levels of escalated conflict. At this point individuals do not respond adaptively to the quality of their partner, instead their preference for a given action is dependent on their own quality.

@@

## Overal effect of variance

In figure \ref{fig:learning_all_var} we show how the effect of increasing 
variance in quality further deacrases the amount of individuals in a population 
expressing the hawk phenotype.

@@

Figures \ref{fig:learning_var0.1} and \ref{fig:learning_all_var} only show how learners behave at a population level, but they do not show how inter-invidivual variation behind those frequencies. In figure \ref{fig:AC_HD} we show the result of the function aproximation mechanism arising at the end of a generation for a sample of the individuals of the populations. These results are show under three different levels of variation in quality. All three panes show that through their learning processes individuals split up in two groups, one which mostly behaves like a hawk and a second that mostly behaves like a dove. The group of "hawks" is mostly, but no exclusively, composed of individuals whose quality is on the upper half of the quality distribution. Concordantly, the group of "doves" is mostly composed of individuals whose qualiy is on the lower half of the quality distribution. The relation between quality and the prefered behaviour of individuals obviously only holds in scenarios with variance in quality. In the scenario with quality invariance, the split of individuals in two groups with a tenedency for each behaviour nevertheless remains.

```{r fig6, echo=FALSE, fig.cap="\\label{fig:AC_HD}Within population variation in the learning parameters at the end of learning for three scenarios of variation in the quality of individuals.", out.width = '80%'}
knitr::include_graphics("Simulations//QualStDv_//WeightsVarQualSt.png")
```

@@

In figure \ref{fig:learn_dyn} we show the learning trajectory, for the actor and the critic, of a sample of individuals. These dynamics show how individuals split in the two groups according to their experiences. Despite the robustness of the bimodal distribution along the learning dynamics, it is not uncommon that individual go from one group to the other.

@@

Finally, in a new set of simulations we explore the question of how competitive learners are in competition with the pure strategies. Figure \ref{fig:types_comp}
show the results of competition among the three strategies. The population starts with a 1:1 ratio of pure hawks anf doves, and learners are intruced by mutation. Learners manage to increase in frequency. They, however, do not manage to outcompete the pure streategies. In the end, there is a polymorphic equilibrium with the three strategies coexisting, with learners being the most frequen of the threee.

>>>>

---
title: "The evolution of badges of status with learners"
author: "Andrés Quiñones"
bibliography: SocialMorphCue.bib
output: 
  pdf_document: 
    fig_caption: yes
    keep_tex: yes
header-includes: \usepackage{float}
                #\floatplacement{figure}{H}
---

@@

Individuals have one of two genetically determined phenotypic strategies. _Hawks_ are willing to start a conflict over resources, while _doves_ prefer to stand down in the hope to share the resource without an aggressive contest. Individuals interact randomly with each other over their lifetimes and collect resources depending on the strategy of theirs and their partners. In contrast to the classical _Hawk-dove_, in here when two _hawks_ interact with each other one of them wins the contest and gets $V-\frac{C}{2}$ as payoff; while the loser gets $-\frac{C}{2}$. Where $V$ is the value of the resource individuals are competing for, and $C$ is the cost of an aggressive contest. individuals share the cost, but only the winner gets the value of the resource.The rest of payoff matrix follows the standard game.

@@

where $Q_A$ and $Q_b$ are the qualities of individuals A and B respectively, and $\beta$ is how important quality is in determining who wins the contest. 
Individuals' reproductive success is proportional to the payoff accumulated
throughout their life. Figure \ref{fig:HD_game} shows the evolutionary dynamics 
of  the two genotypes, which fit the game theoretical prediction.

@@

To complete the receiver side of the signalling system, we introduce a new 
strategy in the game that we call _learner_. _Learners_ use a mixed strategy in 
the hawk-dove game. Which means, they play hawk (dove) with a certain probability.
Furthermore, the probability with which the play each strategy changes during their lifetime according to the payoff they receive in previous interactions. In other words, _learners_ learn to behave in the game. To implement the learning
process, we use the reinforcement learning formalism. Specifically, we use the actor-critic algorithm [@sutton_reinforcement_2018]. Individuals ( _learners_ ) perceive the _badge_ of their counterpart as the environmental state in which they find themselves. They estimate the value of that state and a preference for playing hawk (dove) in that state. They update their estimation and preference after each interaction they have. Given that the badge size is a real number between 0 and 1, there are infinitely many environmental states. Thus, for their value estimation and preference computation, individual must use a form of generalization. We let them
use as generalization a function approximation method based on radial basis functions. Each individual uses five _centers_ spread uniformly throughout the range of badge size variation (from 0 to 1). Each time an individual interacts with a con-specific the size of the badge of the interacting partner triggers a response from each one of the _centers_. This response decreases in strength as the badge size is further from the _center_. The decrease follows a radial basis function with Gaussian shape. The sum of the responses given by all the centers provide the value estimation and the preference for one of the two pure strategies (hawk or dove). During the learning process individuals update the _weight_ of the response that each one of these centers triggers. In the case of the preference for the strategies the numerical value resulting from the sum of the responses is translated to a probability of one of the two pure strategies through a logistic function. Figure \ref{fig:learning_cartoonRBF} shows a made up example of how the function approximation works for the actor (top panel - probability of choosing dove) and the critic (bottom panel - expected payoff from facing an individual with a given badge size).

@@

In Figure \ref{fig:learning_invar} we show the phenotypic frequencies and interaction-type frequencies in 4 generations of a population composed of only _learners_ that do not vary in their quality. The population is also monomorphic for the reaction norm parameters ($\alpha=0$ and $\gamma=0$). With the values chosen the badge size distribution is the same as the quality distribution. Thus, the four generations represent just replicates of the learning process. The results show that under quality invariance, the learning algorithm develops similar phenotypic frequencies as the evolutionary dynamics of the pure strategies.

@@

In figure \ref{fig:learning_var0.1} we show results of phenotypic and interaction types frequencies from simulations in which individuals vary in their quality. 
The value of quality is drawn from a truncated normal distribution with mean 0.5 
and standard deviation 0.1. Compared to the expectation from evolutionary dynamics of the standard game, the learning scenario with quality variation shows lower levels of hawks and therefore lower levels of escalated conflict. At this point individuals do not respond adaptively to the quality of their partner, instead their preference for a given action is dependent on their own quality.

@@

## Overall effect of variance

In figure \ref{fig:learning_all_var} we show how the effect of increasing 
variance in quality further decreases the amount of individuals in a population 
expressing the hawk phenotype.

@@

Figures \ref{fig:learning_var0.1} and \ref{fig:learning_all_var} only show how learners behave at a population level, but they do not show the inter-individual variation behind those frequencies. In figure \ref{fig:AC_HD} we show the result of the function approximation mechanism arising at the end of a generation for a sample of the individuals of the populations. These results are shown under three different levels of variation in quality. All three panes show that through their learning processes individuals split up in two groups, one which mostly behaves like a hawk and a second that mostly behaves like a dove. The group of "hawks" is mostly, but no exclusively, composed of individuals whose quality is on the upper half of the quality distribution. In accordance, the group of "doves" is mostly composed of individuals whose quality is on the lower half of the quality distribution. The relation between quality and the preferred behavior of individuals obviously only holds in scenarios with variance in quality. In the scenario with quality invariance, the split of individuals in two groups with a tendency for each behavior nevertheless remains.

```{r fig6, echo=FALSE, fig.cap="\\label{fig:AC_HD}Within population variation in the learning parameters at the end of learning for three scenarios of variation in the quality of individuals.", out.width = '70%'}
knitr::include_graphics("Simulations//QualStDv_//WeightsVarQualSt.png")
```

@@

In figure \ref{fig:learn_dyn} we show the learning trajectory, for the actor and the critic, of a sample of individuals. These dynamics show how individuals split in the two groups according to their experiences. Despite the robustness of the bi-modal distribution along the learning dynamics, it is not uncommon that individual go from one group to the other.

@@

Finally, in a new set of simulations we explore the question of how competitive learners are in competition with the pure strategies. Figure \ref{fig:types_comp}
show the results of competition among the three strategies. The population starts with a 1:1 ratio of pure hawks and doves, and learners are introduced by mutation. Learners manage to increase in frequency. They, however, do not manage to outcompete the pure strategies. In the end, there is a polymorphic equilibrium with the three strategies coexisting, with learners being the most frequent of the three.

### 2019-07-02 QD_report.tex
FInalize QD report. Plot of evol simulations

\floatplacement{figure}{H}

@@

Classical Hawk-Dove game. Individuals have one fo two genetically
determined phenotypic strategies. \emph{Hawks} are willing to start a
conflict over resources, while \emph{doves} prefer to stand down in the
hope to share the resource without an aggressive contest. Individuals
interact randomly with each other over their lifetimes and collect
resources depending on the strategy of theirs and their partners. In
contrast to the classical \emph{Hawk-dove}, in here when to \emph{hawks}
interact with each other one of them wins the contest and gets
\(V-\frac{C}{2}\) as payoff; while the loser gets \(-\frac{C}{2}\).
Where \(V\) is the value of the resource individuals are competing for,
and \(C\) is the cost of an agressive contest. individuals share the
cost, but only the winner gets the value of the resource.The rest of
payoff matrix follows the standard game.

@@

where \(Q_A\) and \(Q_b\) are the qualities of individuals A and B
respectively, and \(\beta\) is how important quality is in determining
who wins the contest. Individuals' reproductive success is proportional
to the payoff accumulated throughout their lifes. Figure
\ref{fig:HD_game} shows the evolutionary dynamics of the two genotypes,
which fit the game theoretical prediction.

\begin{figure}
\includegraphics[width=0.8\linewidth]{Simulations//mutType_//basicHawkDove} \caption{\label{fig:HD_game}Hawk-dove game. Dashed line is the  game theoretical prediction for frequency of hawks.}\label{fig:fig1}
\end{figure}

@@

To complete the receiver side of the signalling system, we introduce a
new strategy in the game that we call \emph{learner}. \emph{Learners}
use a mixed strategy in the hawk-dove game. Which means, they play hawk
(dove) with a certain probability. Furthermore, the probability with
which the play each strategy changes during their lifetime according to
the payoff they recieve in previous interactions. In other words,
\emph{learners} learn to behave in the game. To implement the learning
process, we use the reinforcement learning formalism. Specifically, we
use the actor-critic algorithm. Individuals ( \emph{learners} ) perceive
the \emph{badge} of their counterpart as the environmental state in
which they find themselves. They estimate the value of that state and a
preference for playing hawk (dove) in that state. They update their
estimation and preference after each interaction they have. Given that
the badge size is a real number between 0 and 1, there are infinitely
many environmental states. Thus, for their value estimation and
preference computation, individual must use a form of generalization. We
let them use as generalization a function approximation method based on
radial basis functions. Each individual uses five \emph{centers} spread
uniformly throughout the range of badge size variation (from 0 to 1).
Each time an individual interacts with a conspecific the size of the
badge of the interacting partner triggers are response from each one of
the \emph{centers}. This response decreases in strength as the badge
size is further from the \emph{center}. The decrease follows a radial
basis function with gaussian shape. The sum of the responses given by
all the centers provide the value estimation and the preference for one
of the two pure strategies (hawk or dove). During the learning process
individuals update the response that each one of these centers triggers.
In the case of the preference for the strategies the numerical value
resulting from the sum of the responses is translated to a probability
of one of the two pure strategies through a logistic function. Figure
\ref{fig:learning_cartoonRBF} shows a made up example of how the
function approximation works for the actor (top panel - probability of
choosing dove) and the critic (bottom panel - expected payoff from
facing an individual with a given badge size).

@@

In Figure \ref{fig:learning_invar} we show the phenotypic frequencies
and interaction-type frequencies in 4 generations of a population
composed of only \emph{learners} that do not vary in their quality. The
population is also monomorphic for the reaction norm parameters
(\(\alpha=0\) and \(\beta=0\)). With the values chosen the badge size
distribution is the same as the quality distribution. Thus, the four
generations respresent just replicates of the learning process. The
results show that under quality invariance, the learning algorithm
develops similar phenotypic frequencies as the evolutionary dynamics of
the pure strategies.

@@

@@

\hypertarget{overal-effect-of-variance}{%
\subsection{Overal effect of variance}\label{overal-effect-of-variance}}

@@

\hypertarget{how-does-learning-look-like-in-the-hawk-dove-game}{%
\subsection{How does learning look like in the Hawk-Dove
game}\label{how-does-learning-look-like-in-the-hawk-dove-game}}

\begin{figure}
\includegraphics[width=0.8\linewidth]{Simulations//QualStDv_//WeightsVarQualSt} \caption{\label{fig:AC_HD}Within population variation in the learning parameters at the end of learning for three scenarios of variation in the quality of individuals.}\label{fig:fig6}
\end{figure}

@@

@@

@@

>>>>

@@

Individuals have one of two genetically determined phenotypic
strategies. \emph{Hawks} are willing to start a conflict over resources,
while \emph{doves} prefer to stand down in the hope to share the
resource without an aggressive contest. Individuals interact randomly
with each other over their lifetimes and collect resources depending on
the strategy of theirs and their partners. In contrast to the classical
\emph{Hawk-dove}, in here when two \emph{hawks} interact with each other
one of them wins the contest and gets \(V-\frac{C}{2}\) as payoff; while
the loser gets \(-\frac{C}{2}\). Where \(V\) is the value of the
resource individuals are competing for, and \(C\) is the cost of an
aggressive contest. individuals share the cost, but only the winner gets
the value of the resource.The rest of payoff matrix follows the standard
game.

@@

where \(Q_A\) and \(Q_b\) are the qualities of individuals A and B
respectively, and \(\beta\) is how important quality is in determining
who wins the contest. Individuals' reproductive success is proportional
to the payoff accumulated throughout their life. Figure
\ref{fig:HD_game} shows the evolutionary dynamics of the two genotypes,
which fit the game theoretical prediction.

\begin{figure}
\includegraphics[width=0.8\linewidth]{Simulations/mutType_/basicHawkDove} \caption{\label{fig:HD_game}Hawk-dove game. Dashed line is the  game theoretical prediction for frequency of hawks.}\label{fig:fig1}
\end{figure}

@@

To complete the receiver side of the signalling system, we introduce a
new strategy in the game that we call \emph{learner}. \emph{Learners}
use a mixed strategy in the hawk-dove game. Which means, they play hawk
(dove) with a certain probability. Furthermore, the probability with
which the play each strategy changes during their lifetime according to
the payoff they receive in previous interactions. In other words,
\emph{learners} learn to behave in the game. To implement the learning
process, we use the reinforcement learning formalism. Specifically, we
use the actor-critic algorithm (Sutton and Barto 2018). Individuals (
\emph{learners} ) perceive the \emph{badge} of their counterpart as the
environmental state in which they find themselves. They estimate the
value of that state and a preference for playing hawk (dove) in that
state. They update their estimation and preference after each
interaction they have. Given that the badge size is a real number
between 0 and 1, there are infinitely many environmental states. Thus,
for their value estimation and preference computation, individual must
use a form of generalization. We let them use as generalization a
function approximation method based on radial basis functions. Each
individual uses five \emph{centers} spread uniformly throughout the
range of badge size variation (from 0 to 1). Each time an individual
interacts with a con-specific the size of the badge of the interacting
partner triggers a response from each one of the \emph{centers}. This
response decreases in strength as the badge size is further from the
\emph{center}. The decrease follows a radial basis function with
Gaussian shape. The sum of the responses given by all the centers
provide the value estimation and the preference for one of the two pure
strategies (hawk or dove). During the learning process individuals
update the \emph{weight} of the response that each one of these centers
triggers. In the case of the preference for the strategies the numerical
value resulting from the sum of the responses is translated to a
probability of one of the two pure strategies through a logistic
function. Figure \ref{fig:learning_cartoonRBF} shows a made up example
of how the function approximation works for the actor (top panel -
probability of choosing dove) and the critic (bottom panel - expected
payoff from facing an individual with a given badge size).

@@

In Figure \ref{fig:learning_invar} we show the phenotypic frequencies
and interaction-type frequencies in 4 generations of a population
composed of only \emph{learners} that do not vary in their quality. The
population is also monomorphic for the reaction norm parameters
(\(\alpha=0\) and \(\gamma=0\)). With the values chosen the badge size
distribution is the same as the quality distribution. Thus, the four
generations represent just replicates of the learning process. The
results show that under quality invariance, the learning algorithm
develops similar phenotypic frequencies as the evolutionary dynamics of
the pure strategies.

@@

In figure \ref{fig:learning_var0.1} we show results of phenotypic and
interaction types frequencies from simulations in which individuals vary
in their quality. The value of quality is drawn from a truncated normal
distribution with mean 0.5 and standard deviation 0.1. Compared to the
expectation from evolutionary dynamics of the standard game, the
learning scenario with quality variation shows lower levels of hawks and
therefore lower levels of escalated conflict. At this point individuals
do not respond adaptively to the quality of their partner, instead their
preference for a given action is dependent on their own quality.

@@

\hypertarget{overall-effect-of-variance}{%
\subsection{Overall effect of
variance}\label{overall-effect-of-variance}}

In figure \ref{fig:learning_all_var} we show how the effect of
increasing variance in quality further decreases the amount of
individuals in a population expressing the hawk phenotype.

@@

\hypertarget{how-does-learning-look-like-in-the-hawk-dove-game}{%
\subsection{How does learning look like in the Hawk-Dove
game?}\label{how-does-learning-look-like-in-the-hawk-dove-game}}

Figures \ref{fig:learning_var0.1} and \ref{fig:learning_all_var} only
show how learners behave at a population level, but they do not show the
inter-individual variation behind those frequencies. In figure
\ref{fig:AC_HD} we show the result of the function approximation
mechanism arising at the end of a generation for a sample of the
individuals of the populations. These results are shown under three
different levels of variation in quality. All three panes show that
through their learning processes individuals split up in two groups, one
which mostly behaves like a hawk and a second that mostly behaves like a
dove. The group of ``hawks'' is mostly, but no exclusively, composed of
individuals whose quality is on the upper half of the quality
distribution. In accordance, the group of ``doves'' is mostly composed
of individuals whose quality is on the lower half of the quality
distribution. The relation between quality and the preferred behavior of
individuals obviously only holds in scenarios with variance in quality.
In the scenario with quality invariance, the split of individuals in two
groups with a tendency for each behavior nevertheless remains.

\begin{figure}
\includegraphics[width=0.7\linewidth]{Simulations//QualStDv_//WeightsVarQualSt} \caption{\label{fig:AC_HD}Within population variation in the learning parameters at the end of learning for three scenarios of variation in the quality of individuals.}\label{fig:fig6}
\end{figure}

@@

In figure \ref{fig:learn_dyn} we show the learning trajectory, for the
actor and the critic, of a sample of individuals. These dynamics show
how individuals split in the two groups according to their experiences.
Despite the robustness of the bi-modal distribution along the learning
dynamics, it is not uncommon that individual go from one group to the
other.

@@

Finally, in a new set of simulations we explore the question of how
competitive learners are in competition with the pure strategies. Figure
\ref{fig:types_comp} show the results of competition among the three
strategies. The population starts with a 1:1 ratio of pure hawks and
doves, and learners are introduced by mutation. Learners manage to
increase in frequency. They, however, do not manage to outcompete the
pure strategies. In the end, there is a polymorphic equilibrium with the
three strategies coexisting, with learners being the most frequent of
the three.

@@

\hypertarget{cited-literature}{%
\section*{Cited literature}\label{cited-literature}}
\addcontentsline{toc}{section}{Cited literature}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-sutton_reinforcement_2018}{}%
Sutton, Richard S., and Andrew G. Barto. 2018. \emph{Reinforcement
Learning: An Introduction}. Edited by Francis Bach. Second edition
edition. Cambridge, MA: A Bradford Book.
